---
title: "Personal Dataset Project"
author: "Hannah"
date: "4/17/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
```

## Week 7

What were the most commonly recalled types of food between 2017-2020?
What were the most common reasons foods were recalled?
Was a particular category of food more likely to be recalled? Are conventional or organic foods more likely to be recalled?

## Week 9

### (B) Write a brief paragraph describing why you decided on this data source, where the data comes from originally, how the chosen data will help you answer a question you are interested in (not necessarily one of the questions you wrote about in Week \#7 but hopefully something related), and any processing steps you applied to the data. Submit the paragraph as a response to this question - you do not have to post this to github. 

The data is from the FDA website. They have posted a description of all the recalled food and drug items in the US between 2017-2020. It is an open data source with over a thousand entries. I removed all of the entries related to drugs, cosmetics, medical devices, etc. by narrowing the Type column to Food. I hope to answer questions such as: what is the most common reason food is recalled and are there certain types of foods which are more likely to be recalled? 

Processing steps:

```{r loaddata}
#upload csv as a data frame

recalls <- read.csv("recalls.csv")
head(recalls)

```

```{r processing}

#date column data type

recalls$Date <- as.Date(recalls$Date)

#separate elements of Product.Types column

types <- str_split(recalls$Product.Types,",")
recalls$Category <- sapply(types,"[",1)

recalls <- recalls[recalls$Category == "Food & Beverages",]

types <- str_split(recalls$Product.Types,",")
recalls$Food.Sub <- sapply(types,tail,1)

recalls$Food.Sub <- str_trim(recalls$Food.Sub)

#replace Food & Beverage Safety

recalls$Food.Sub[recalls$Food.Sub == "Food & Beverage Safety"] <- ""
recalls$Food.Sub[recalls$Food.Sub == "Allergens"] <- ""
recalls$Food.Sub[recalls$Food.Sub == "Food & Beverages"] <- ""
recalls$Food.Sub[recalls$Food.Sub == "Foodborne Illness"] <- ""

head(recalls)

write.csv(recalls, "recalls_clean.csv", row.names = FALSE)
```

## Week 11

```{r cleaning continued}
rc <- read.csv("recalls_clean.csv")

#there are multiple csv files because I worked on this project over the course of several weeks. 

rc$Recall.Reason.Description <- str_to_title(rc$Recall.Reason.Description)
head(rc)

#string data cleaning

rc$Recall.Reason.Description[rc$Recall.Reason.Description=="Undeclared Peanut"] <- "Undeclared Peanuts"
rc$Recall.Reason.Description[rc$Recall.Reason.Description=="Undeclared Almond"] <- "Undeclared Almonds"
rc$Recall.Reason.Description[rc$Recall.Reason.Description=="Potential To Be Contaminated With Listeria Monocytogenes"] <- "Potential To Be Contaminated With Listeria"
rc$Recall.Reason.Description[rc$Recall.Reason.Description=="Listeria Monocytogenes"] <- "Listeria"
rc$Recall.Reason.Description[rc$Recall.Reason.Description=="Allergen - Eggs"] <- "Allergen: Eggs"
rc$Recall.Reason.Description[rc$Recall.Reason.Description=="Allergen: Egg"] <- "Allergen: Eggs"
rc$Recall.Reason.Description[rc$Recall.Reason.Description=="Undeclared Egg"] <- "Undeclared Eggs"

write.csv(rc, "rc.csv", row.names = FALSE)
```

```{r reason df}

#create new data frame to splice out "Recall Reason Description" string values that commonly appear. 

reason <- data.frame(table(rc$Recall.Reason.Description))
 names(reason)[names(reason) == "Var1"] <- "Reason.Description"
head(reason)
reason[reason$Freq >10,]

#create new object only with the string values of interest (common "Recall Reason Description" values)

reason2 <- rc[
  rc$Recall.Reason.Description == "Undeclared Milk" |
  rc$Recall.Reason.Description == "Listeria" |  
  rc$Recall.Reason.Description == "Potential To Be Contaminated With Listeria" |
  rc$Recall.Reason.Description == "Undeclared Eggs" |
  rc$Recall.Reason.Description == "Salmonella" |
  rc$Recall.Reason.Description == "Undeclared Peanuts" |
  rc$Recall.Reason.Description == "Potential To Be Contaminated With Salmonella" |
  rc$Recall.Reason.Description == "Undeclared Sulfites" |
  rc$Recall.Reason.Description == "Undeclared Almonds" |
  rc$Recall.Reason.Description == "Undeclared Soy"
    ,]

head(reason2)
```

```{r reason2 plot}

#create a visualization showing the frequencies of the most common "Recall Reason Descriptions"

ggplot(reason2, aes(x=Recall.Reason.Description, color=Recall.Reason.Description)) +
  geom_bar() +
  xlab("Reason") + ylab("Occurances") +
  ggtitle("Common Reasons for Recalled Food") +
  coord_flip() +
  theme(legend.position = "none")
```

## Upload to githut
```{r save plot}

##export plot to upload to github

png("recall.reasons.png")
recall.reasons <- ggplot(reason2, aes(x=Recall.Reason.Description, color=Recall.Reason.Description)) +
  geom_bar() +
  xlab("Reason") + ylab("Occurances") +
  ggtitle("Common Reasons for Recalled Food") +
  coord_flip() +
  theme(legend.position = "none")
print(recall.reasons)
```

## EDA
```{r EDA}

recall.category <- data.frame(table(reason2$Recall.Reason.Description, reason2$Food.Sub))
  names(recall.category)[names(recall.category) == "Var1"] <- "Reason.Description"
  names(recall.category)[names(recall.category) == "Var2"] <- "Food.Subcategory"
  recall.category[recall.category$Freq>5,]
```

## EDA Reasoning
Data frame created from a contingency table to show the most common associations between recall reason and food group category. For example undeclared milk is frequently associated with processed convenience foods including: bakery products, chocolate products and snack food items. Also seen in this table is that the underlying dataset is not as clean as I would like it to be. "Food-borne Illness" is not a category of food but for some reason it is showing up in that column. There are also a bunch of entries which don't have anything listed in the Food.Subcategory column because nothing was listed in the original dataset. 

## create contingency table visualization to export
## create food.sub visualizations of most common
```{r foodsub df}

#create new data frame to splice out Food Subcategory string values that commonly appear. 

foodsub <- data.frame(table(rc$Food.Sub))
 names(foodsub)[names(reason) == "Var1"] <- "Food Subcategory"
head(foodsub)
foodsub[foodsub$Freq >10,]

#create new object only with the string values of interest (common Food Subcategory values)


foodsub2 <- rc[
  rc$Food.Sub == "Bakery Product/Mix" |
  rc$Food.Sub == "Chocolate/Cocoa Product" |  
  rc$Food.Sub == "Cheese/Cheese Product" |
  rc$Food.Sub == "Fruit/Fruit Product" |
  rc$Food.Sub == "Flavors &  Salts" |
  rc$Food.Sub == "Fish" |
  rc$Food.Sub == "Ice Cream/Frozen Dairy" |
  rc$Food.Sub == "Gravy/Sauces" |
  rc$Food.Sub == "Snack Food Item" |
  rc$Food.Sub == "Produce" |
  rc$Food.Sub == "Prepared Food" |
  rc$Food.Sub == "Meal Replacements" |
  rc$Food.Sub == "Milk/Milk Product"
    ,]

head(foodsub2)
```

```{r foodsub2 plot}

#create a visualization showing the frequencies of the most common Food Subcategories

ggplot(foodsub2, aes(x=Food.Sub, color=Food.Sub)) +
  geom_bar() +
  xlab("Food Subcategory") + ylab("Occurances") +
  ggtitle("Common Food Categories of Recalled Food") +
  coord_flip() +
  theme(legend.position = "none")
```

```{r save plot2}

##export plot to upload to github

png("foodsub.plot.png")
foodsub.plot <- ggplot(foodsub2, aes(x=Food.Sub, color=Food.Sub)) +
  geom_bar() +
  xlab("Food Subcategory") + ylab("Occurances") +
  ggtitle("Common Food Categories of Recalled Food") +
  coord_flip() +
  theme(legend.position = "none")
print(foodsub.plot)
```

```{r interestingreasons}

unique(rc$Recall.Reason.Description, incomparables = FALSE)

"Presence Of Small Pieces Of Plastic"  
"May Be Contaminated With Rodent Droppings"
"Misbranding & Unapproved New Drugs"
"Unsafe Levels Of Arsenic And Lead"                                                                                          
"Foreign Object"                                                                                                             
"Potential To Be Contaminated With Hepatitis A"                                                                              
"Potential Presence Of Plastic And Metal"

#Use excel to create table

```

```{r brand df}

#create new data frame to splice out Brand Name string values that commonly appear. 

brand <- data.frame(table(rc$Brand.Names))
 names(brand)[names(brand) == "Var1"] <- "Brand.Names"
head(brand)
brand[brand$Freq >4,]

#use excel to create table

```

```